---
title: "GCCR004 Data Analysis - Part 1"
author: "Ha Nguyen"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output:
  pdf_document: default
  html_notebook: default
Note: Updated data analysis in November 2021; 
  (1) Multivariate analysis
  (2) Hierarchical Cluster Analysis
  (3) Experienced intensities by time period since COVID onset symptoms
  Run this part after you run Part 0 of the data analyses.
---
```{r, echo=FALSE, results='hide', warning=FALSE, message=FALSE}
# load packages
Packages <- c("tidyverse", "Hmisc", "FactoMineR", "questionr", "cluster", "lme4","nlme","multcomp","ggpubr","MuMIn","effects","LMERConvenienceFunctions","lsmeans",
              "datawrangling","factoextra","plotly","yarrr","cowplot","ggExtra","sjstats", "car")
lapply(Packages, library, character.only = TRUE)
```

# 1. Multivariate analysis for 6 original groups - MFA with centered data 
```{r, echo=FALSE, warning=FALSE, message=FALSE}
library(FactoMineR)
#Removing 6-month ratings
mfa_today<-MFA(dplyr::select(dat,c(group,
                                   Smell_ability,Taste_ability,Oralirritation_ability,
                            Cosmetics,Spices,Fruits,Other_items,
                            Sweet,Salty,Sour,Bitter,Nasalirritation,Oralirritation)),
               group = c(1,3,4,4,1,1),type=c("n",rep("c",5)),
               name.group = c("Group","Self-reported abilities","Smell intensities",
                              "Taste intensities","Nasalirritation intensity",
                              "Oralirritation intensity"),
               num.group.sup = c(1,2),graph=F)
```
## Figure 3B - Correlation circle
```{r, echo=FALSE, warning=FALSE, message=FALSE,fig.width=10}
library(ggpubr)
plot(mfa_today,choix="var",graph.type = "ggplot")+
  scale_color_manual(labels=c("Smell intensities","Taste intensities","Nasalirritation intensity",
                              "Oralirritation intensity","Self-reported abilities"),
                     values=c("#638FBB","#EC922D","#7f462c","#178F8C","#736f6e"))+
  theme_classic()+
  border()
ggsave("figures/Figure 3A_MFA_correlation circle_2.tiff",device="tiff",dpi = 300)
```

## Figure 3A - Group plot
Color codes for 6 groups: c("#bd0000","#f19d83","#252538","#5e5d75","#9d9bb7","#da6142")
```{r, echo=FALSE, warning=FALSE, message=FALSE,fig.width=7}
#library(ggpubr)
plotellipses(mfa_today,keepvar=1,invisible="ind",graph.type = "ggplot",
             palette=c("#bd0000","#f19d83","#252538","#5e5d75","#9d9bb7","#da6142"))+
  theme_classic()+
  border()
ggsave("figures/Figure 3B_MFA_6groups_4.tiff",device="tiff",dpi = 300)
```

# 2. Hierarchical Cluster Analysis
## HCA - Grouping participants based on taste, smell, and oralirritation intensities
```{r, echo=FALSE, warning=FALSE, message=FALSE}
dat_clust<-dat%>%drop_na(Taste_ability,Smell_ability,Oralirritation_ability,Taste_intensity,Smell_intensity,Oralirritation) # remove any missing values in this dataset, returning 8319 obs.
```

```{r, echo=FALSE, warning=FALSE, message=FALSE}
library(cluster)
library(factoextra)
fviz_nbclust(dat_clust[,c("Taste_intensity","Smell_intensity","Oralirritation")],FUN=hcut,method="wss",k.max=10)
```

The optimal number of clusters is 3
```{r, echo=FALSE, warning=FALSE, message=FALSE}
set.seed(123)
hc <- agnes(dat_clust[,c("Taste_intensity","Smell_intensity","Oralirritation")], method = "ward")
sup_grp<-cutree(hc,k=3)
dat_hc<-cbind(dat_clust,sup_grp)
dat_hc$sup_grp<-as.factor(dat_hc$sup_grp)
```

## Figure 4A - 3D plots for 3 clusters 
```{r, echo=FALSE, warning=FALSE, message=FALSE}
library(plotly)
plot_ly(dat_hc,x=~Smell_intensity,y=~Taste_intensity,
           z=~Oralirritation,color=~sup_grp,
           colors =c("#009E73","#D55E00","#56B4E9"))%>%
         layout(legend=list(title=list(text='<b> Cluster </b>')),
                scene=list(xaxis=list(title="Smell intensity"),
                           yaxis=list(title="Taste intensity"),
                           zaxis=list(title="Chemesthesis intensity")
                           )
                )%>%
  add_markers(size=2)
```

## Figure 4B - Pirate pLots for 3 clusters 
```{r, echo=FALSE, warning=FALSE, message=FALSE}
# Ability by modality and cluster
dat_ability<-dat_hc%>%
  pivot_longer(c(Smell_ability,Taste_ability,Oralirritation_ability),names_to = "Modality",values_to = "Ability")
dat_ability$Modality <- fct_recode(dat_ability$Modality, "Smell" = 'Smell_ability', "Taste" = 'Taste_ability',"Oralirritation"="Oralirritation_ability")
dat_ability$Modality<-fct_relevel(dat_ability$Modality,"Smell","Taste","Oralirritation")

library(yarrr)

pdf(file = "figures/clust_ability.pdf", width = 12, height = 6) 

pirateplot(formula = Ability ~ Modality + sup_grp, 
data = dat_ability,
theme = 2,
main = "",
ylab = "", yaxt = "n", # remove Y axis to add a new one after
ylim = c(0, 100), 
xlab = "", xaxt = "n",
bean.b.o = 1, # borde violin
bean.f.col = "white",
bean.b.col = "black",
jitter.val = 0.05, # dispersión de los puntos
point.o = 0.15, # point opacity 
point.pch = 20, # forma de los puntos
point.col = c("#009E73","#009E73","#009E73",
              "#D95F02","#D95F02","#D95F02",
              "#56B4E9","#56B4E9","#56B4E9"), # point color by group
point.cex = 0.5,
#bar.f.o = 0.5, bar.b.o = 1, # barplot
gl.col = "white", #  gridlines
avg.line.fun = median,
inf.method = "iqr", # rango interquartil
inf.disp = "bean",
inf.f.col = c("#009E73","#009E73","#009E73",
              "#D95F02","#D95F02","#D95F02",
              "#56B4E9","#56B4E9","#56B4E9"))

#mtext("A", side=3, outer=T, adj=0.0, line=-1.5, font=2, cex=2)
axis(2, at = seq(from = 0, to = 100, by = 20), cex.axis=1.5)  # new y axis with less breaks
axis(1, at=c(1,2,3,5,6,7,9,10,11),labels=c("Smell","Taste","Chemesthesis","Smell","Taste","Chemesthesis","Smell","Taste","Chemesthesis"), tck=-0.01, cex.axis=1) 
axis(3, at=c(2,6,10), labels=c("Cluster 1", "Cluster 2", "Cluster 3"),
     tck=0, cex.axis=1.5, font=2)
title(ylab = "Ability", col.lab ="black", cex.lab=1)

dev.off()
```

```{r, echo=FALSE, warning=FALSE, message=FALSE}
# Perceived intensity by modality and cluster
dat_intensity<-dat_hc %>%
  pivot_longer(c(Smell_intensity,Taste_intensity,Oralirritation),names_to = "Modality",values_to = "Intensity")
dat_intensity$Modality <- fct_recode(dat_intensity$Modality, "Smell" = 'Smell_intensity', "Taste" = 'Taste_intensity',"Oralirritation"="Oralirritation")
dat_intensity$Modality<-fct_relevel(dat_intensity$Modality,"Smell","Taste","Oralirritation")
#library(yarrr)
pdf(file = "figures/clust_intensity.pdf", width = 12, height = 6) 

pirateplot(formula = Intensity ~ Modality + sup_grp, 
data = dat_intensity,
theme = 2,
main = "",
ylab = "", yaxt = "n", # remove Y axis to add a new one after
ylim = c(0, 100), 
xlab = "", xaxt = "n",
bean.b.o = 1, # borde violin
bean.f.col = "white",
bean.b.col = "black",
jitter.val = 0.05, # dispersión de los puntos
point.o = 0.15, # point opacity 
point.pch = 20, # forma de los puntos
point.col = c("#009E73","#009E73","#009E73",
              "#D95F02","#D95F02","#D95F02",
              "#56B4E9","#56B4E9","#56B4E9"), # point color by group
point.cex = 0.5,
#bar.f.o = 0.5, bar.b.o = 1, # barplot
gl.col = "white", #  gridlines
avg.line.fun = median,
inf.method = "iqr", # rango interquartil
inf.disp = "bean",
inf.f.col = c("#009E73","#009E73","#009E73",
              "#D95F02","#D95F02","#D95F02",
              "#56B4E9","#56B4E9","#56B4E9")) +

#mtext("A", side=3, outer=T, adj=0.0, line=-1.5, font=2, cex=2)
axis(2, at = seq(from = 0, to = 100, by = 20), cex.axis=1.5)  # new y axis with less breaks
axis(1, at=c(1,2,3,5,6,7,9,10,11),labels=c("Smell","Taste","Chemesthesis","Smell","Taste","Chemesthesis","Smell","Taste","Chemesthesis"), tck=-0.01, cex.axis=1) 
axis(3, at=c(2,6,10), labels=c("Cluster 1", "Cluster 2", "Cluster 3"),tck=0, cex.axis=1.5, font=2) 
title(ylab = "Intensity", col.lab ="black", cex.lab=1)

dev.off()
```

## Supp. Table 5. Description of 3 clusters by demographics
Subject characteristics for 3 clusters identified by the Agglomerative Hierarchical Clustering (AHC) on perceived intensities
```{r, echo=FALSE, warning=FALSE, message=FALSE}
summary(dat_hc$sup_grp)

# gender
table(dat_hc$gender,dat_hc$sup_grp)
# addmargins(table(dat_hc$gender,dat_hc$sup_grp)) # add sum
round(100*prop.table(table(dat_hc$gender,dat_hc$sup_grp),2),digits=1) # column proportion
chisq.test(table(dat_hc$gender,dat_hc$sup_grp)[c("Female","Male"),])

# age
dat_hc %>%
  group_by(sup_grp)%>%
  summarise(n=sum(!is.na(age)),mean=mean(age,na.rm=T),sd=sd(age,na.rm = T))
summary(aov(age ~ sup_grp, dat_hc))
summary(glht(aov(age ~ sup_grp, dat_hc), linfct = mcp(sup_grp = "Tukey")))

# period_covid
dat_hc %>%
  group_by(sup_grp)%>%
  summarise(n=sum(!is.na(period_covid)),mean=mean(period_covid,na.rm=T),sd=sd(period_covid,na.rm = T))
summary(aov(period_covid ~ sup_grp, dat_hc))
summary(glht(aov(period_covid ~ sup_grp, dat_hc), linfct = mcp(sup_grp = "Tukey")))

# education_rec
table(dat_hc$education_rec,dat_hc$sup_grp)
round(100*prop.table(table(dat_hc$education_rec,dat_hc$sup_grp),2),digits=1) # column proportion
chisq.test(table(dat_hc$education_rec,dat_hc$sup_grp))

# ses
table(dat_hc$ses,dat_hc$sup_grp)
round(100*prop.table(table(dat_hc$ses,dat_hc$sup_grp),2),digits=1) # column proportion
#chisq.test(table(dat_hc$ses,dat_hc$sup_grp))
chisq.test(table(dat_hc$ses,dat_hc$sup_grp)[c("Upper","Upper middle","Lower middle","Lower"),]) # not including "Prefer not to say"

# smoker
table(dat_hc$smoker,dat_hc$sup_grp)
round(100*prop.table(table(dat_hc$smoker,dat_hc$sup_grp),2),digits=1) # column proportion
#chisq.test(table(dat_hc$smoker,dat_hc$sup_grp))
chisq.test(table(dat_hc$smoker,dat_hc$sup_grp)[c("No","Yes - current smoker","Yes - former smoker"),]) # # not including "Prefer not to say"
```

# 3. Perceived intensities by time period since COVID onset symptoms
Recode 'period' into 3 groups: 0-7, 8-14, 15+ (period_covid3)
```{r, echo=FALSE, warning=FALSE, message=FALSE}
## Cutting dat$period into dat_s$period_rec3
dat$period_rec3 <- cut(dat$period_covid,
  include.lowest = FALSE,
  right = FALSE,
  dig.lab = 2,
  breaks = c(0, 8, 15, 393),
  labels = c("0-7","8-14","15+")
)
```

## Supp. Figure 2. Intensity by period_rec3 for COVID+ 
Perceived intensities for smell, taste, and irritation in participants who completed the survey within the first 7 days (n=737), 8-14 days (n=597), and >15 days (n=1,831) of the onset of COVID-19 illness.
```{r, echo=FALSE, warning=FALSE, message=FALSE}
#Intensity by modality and period_rec3
intensity_period3<-dat[dat$group=="COVID"&!is.na(dat$period_rec3),] %>%
  pivot_longer(c(Smell_intensity,Taste_intensity,Oralirritation),names_to = "Modality",values_to = "Intensity")
intensity_period3$Modality <- fct_recode(intensity_period3$Modality, "Smell" = 'Smell_intensity', "Taste" = 'Taste_intensity',"Oralirritation"="Oralirritation")
intensity_period3$Modality<-fct_relevel(intensity_period3$Modality,"Smell","Taste","Oralirritation")
#library(yarrr)
pdf(file = "figures/intensity_by_COVIDperiod3.pdf", width = 12, height = 6) 

pirateplot(formula = Intensity ~ Modality + period_rec3, 
data = intensity_period3,
theme = 2,
main = "",
ylab = "", yaxt = "n", # remove Y axis to add a new one after
ylim = c(0, 100), 
xlab = "", xaxt = "n",
bean.b.o = 1, # borde violin
bean.f.col = "white",
bean.b.col = "black",
jitter.val = 0.05, # dispersión de los puntos
point.o = 0.15, # point opacity 
point.pch = 20, # forma de los puntos
point.col = c("#638FBB","#EC922D","#178F8C"), # point color by modality
point.cex = 0.5,
#bar.f.o = 0.5, bar.b.o = 1, # barplot
gl.col = "white", #  gridlines
avg.line.fun = median,
inf.method = "iqr", # rango interquartil
inf.disp = "bean",
inf.f.col = c("#638FBB","#EC922D","#178F8C")) +

#mtext("A", side=3, outer=T, adj=0.0, line=-1.5, font=2, cex=2)
axis(2, at = seq(from = 0, to = 100, by = 20), cex.axis=1.5)  # new y axis with less breaks
axis(1, at=c(1,2,3,5,6,7,9,10,11),labels=c("Smell","Taste","Chemesthesis","Smell","Taste","Chemesthesis","Smell","Taste","Chemesthesis"), tck=-0.01, cex.axis=1) 
axis(3, at=c(2,6,10), labels=c("0-7 days","8-14 days","15+ days"), #(n=737) (n=597) (n=1831)
     tck=0, cex.axis=1.5, font=2)
title(ylab = "Intensity", col.lab ="black", cex.lab=1)
#"#009E73","#D95F02","#56B4E9"
dev.off()
```

## Supp. Table 4. ANOVAs for period_rec3 for COVID+ (n=3165)
Posthoc tests for perceived intensities for smell, taste, and irritation in participants who completed the survey within the first 7 days (n=737), 8-14 days (n=597), and >15 days (n=1,831) of the onset of COVID-19 illness.
```{r}
# Smell_intensity
library(car)
leveneTest(Smell_intensity ~ period_rec3, data = dat[dat$group=="COVID"&!is.na(dat$period_rec3),]) 

AnovaSmell3 <- aov(Smell_intensity ~ period_rec3, data = dat[dat$group=="COVID"&!is.na(dat$period_rec3),])
summary(AnovaSmell3)
# Homogeneity of variances
plot(AnovaSmell3, 1)

library(multcomp)
summary(glht(AnovaSmell3, linfct = mcp(period_rec3 = "Tukey")))

#library(sjstats)
effectsize::eta_squared(AnovaSmell3, partial = TRUE,alternative = "two")
```

```{r}
# Taste_intensity
#library(car)
leveneTest(Taste_intensity ~ period_rec3, data = dat[dat$group=="COVID"&!is.na(dat$period_rec3),]) 

AnovaTaste3 <- aov(Taste_intensity ~ period_rec3, data = dat[dat$group=="COVID"&!is.na(dat$period_rec3),])
summary(AnovaTaste3)
# Homogeneity of variances
plot(AnovaTaste3, 1)

#library(multcomp)
summary(glht(AnovaTaste3, linfct = mcp(period_rec3 = "Tukey")))

#library(sjstats)
effectsize::eta_squared(AnovaTaste3, partial = TRUE,alternative = "two")
```

```{r}
# Oralirritation
#library(car)
leveneTest(Oralirritation ~ period_rec3, data = dat[dat$group=="COVID"&!is.na(dat$period_rec3),]) 

#AnovaOral3 <- aov(Oralirritation ~ period_rec3, data = dat[dat$group=="COVID"&!is.na(dat$period_rec3),])
summary(AnovaOral3)
# Homogeneity of variances
plot(AnovaOral3, 1)

#library(multcomp)
summary(glht(AnovaOral3, linfct = mcp(period_rec3 = "Tukey")))

#library(sjstats)
effectsize::eta_squared(AnovaOral3, partial = TRUE,alternative = "two")
```

